Brownian motion of main-belt asteroids on human timescales

### 2025-03-03 main.tex
Fix A table for dex, dey
For PSJ submission

\begin{abstract}
  The explosion of high-precision astrometric data on main-belt
  asteroids (MBAs) opens the door to their use for many possible
  inferences on gravitational and non-gravitational forces present in
  this region.  We estimate the size of positional shifts on MBAs
  caused by mutual gravitational encounters with other MBAs that are
  either omitted from ephemeris models or have uncertain mass
  estimates. In other words, what is the typical Brownian motion among
  MBAs that cannot be predicted from the ephemeris, and therefore
  serves as noise on inferences from the MBAs?  We calculate the RMS
  azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by
  numerical estimation of the distribution of impulse sizes and
  directions among known MBAs, combined with analytical propagation
  into future positional uncertainties.  The result depends strongly
  on how well the masses of the larger MBAs are known.  At current
  level of asteroid-mass knowledge, $\sigma_\phi$ rises to
  $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
  $T^{3/2}.$  This is large enough to degrade many inferences from
  Gaia and LSST MBA data, although the LSST data will also improve MBA
  mass knowledge enough to lower this Brownian uncertainty by
  $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are
  suppressed by factors $\approx7$ and $\approx45,$ respectively,
  compared to the azimuthal signal after 10~yr, and grow at $T^{3/2}$
  and $T^{1/2}.$  We suggest that for full exploitation of Gaia and
  LSST MBA data, ephemeris models should include the
  $\approx1000$~largest asteroids as active bodies with free masses,
  even if not all are well constrained.  This will correctly propagate
  the uncertainties from these 1000 sources' deflections into desired
  inferences. The RMS value of deflections from less-massive MBAs is limited to $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

\keywords{asteroids}

@@

We now have all the information we need to create the $\matA$ matrix.
Rotating to $\vecq=\{a, \tau, e_x, e_y, L_x, L_y\}$ and with $\vecI=\{I_r, I_\phi, I_z\}$ we have, to $O(e)$,
\begin{equation}
         \matA(e,T,t_i) = \left( \begin{array}{ccc}
 2e\sin\tau & 2 + 2e\cos\tau & 0 \\
 -2+e\left[1.5\cos\tau-3(T-t_i)\sin\tau\right] & e\sin\tau -3(1+e\cos\tau)(T-t_i) & 0\\
 -\cos\tau-2e\sin^2\tau & 2\cos\tau - 1.5e\sin 2\tau & 0\\
  -\sin\tau + e\sin 2\tau & 2\sin\tau-3e\cos^2\tau & 0 \\
0 & 0 & \sin\tau + 0.5 e \sin 2\tau \\
 0 & 0 &  -\cos\tau +e(3-\cos 2\tau)/2 \\
 \end{array}\right)
\end{equation}
We can now do the time integral and sum over impulse directions in \eqq{eq:Cqjk}, and average over $\tau$ to average over the orbital phase of the impulses. The diagonal elements of $\covm^q$ are
\begin{eqnarray}
  \Var(a) & = & 4T \nIphi \nonumber\\
  \Var(\tau) & = & 4T\nIr + 3T^3\nIphi \nonumber\\
  \Var(e_x) =  \Var(e_y) & = & \nIr/2+2\nIphi \nonumber\\
  \Var( L_x) =  \Var( L_y) & = & \nIz/2
\label{eq:Cq}
\end{eqnarray}
All of the off-diagonal terms (covariances) are either zero or $O\left(e\left\langle nI^2\right\rangle\right),$ and they result in observable consequences that are smaller by $\left\langle e^2 \right\rangle$ than the diagonal terms' contributions.

% \begin{equation}
%   \matC^2 = \left(\begin{array}{cccccc}
%     4T \nIphi & -3T^2\nIphi & e\left(\nIr + 5\nIphi\right) & e\nIr & 0 & 0 \\
% -3T^2\nIphi & 4\nIr + 3T^3\nIphi & -e(11T\nIr/4 + 15T^2\nIphi/4) & e\left[(2T-3T^2/4)\nIr+(-T+9T^2/4)\nIphi\right] & 0 & 0\\
%  e\left(\nIr + 5\nIphi\right) & -e(11T\nIr/4 + 15T^2\nIphi/4) & \nIr/2+2\nIphi & 0 & 0 & 0 \\
%  e\nIr &  e\left[(2T-3T^2/4)\nIr+(-T+9T^2/4)\nIphi\right]  & 0 & \nIr/2+2\nIphi & 0 & 0 \\
% 0 & 0 & 0 & 0 & \nIz/2 & 0 \\
% 0 & 0 & 0 & 0 & 0 \nIz/2 \\
%\end{array}\right)
% \end{equation}
 \subsection{Astrometric and ranging shifts from elements shifts}
\label{sec:observe}
The radial coordinate of a Keplerian orbit is $r=a(1-e\cos E),$ and an expansion of the eccentric anomaly $E$ in powers of $e$ is
\begin{equation}
  \label{eq:Ee2}
  e \cos E  =  e\cos \ma - (e \sin \ma)^2 + O(e^3).
\end{equation}
From this we can derive the perturbations to $r$ as a function of our chosen basis for perturbations to the orbital elements.  Making use of Equations~(\ref{eq:ecosM}) and (\ref{eq:esinM}) we obtain
\begin{equation}
  \Delta r = (1-e\cos\tau) \Delta a + e\sin\tau \Delta\tau +  (-\cos\tau -2e\sin^2\tau) \Delta e_x + (-\sin\tau+e\sin 2\tau) \Delta e_y.
\end{equation}
If we square this expression, average over the orbital phase $\tau$ of the observation, and retain leading order in $e$, we get the variance of the radial coordinate from the unperturbed orbit.  All of the terms involving covariances between orbital elements either average to zero over an orbit, or are suppressed by $e^2$ relative to other terms.  We are left with
\begin{eqnarray}
  \sigma^2_r = \Var(r) & \approx & \Var(a) + \frac{1}{2}\Var(e_x) + \frac{1}{2}\Var(e_y) + \frac{\langle e^2\rangle}{2} \Var(\tau) \\
          & \approx & \frac{T}{2} \nIr + \left( 6T + \frac{3\langle e^2 \rangle}{2} T^3\right) \nIphi.
                      \label{eq:varr}
\end{eqnarray}
The second line substitutes the results from Equations~(\ref{eq:Cq}).  We retain the final term because, as $T>>1,$ this precession term that scales with $\langle e^2 \rangle T^3$ becomes comparable to or larger than the other terms that scale as $T$.  
The heliocentric azimuthal angle $\phi$ is given by \eqq{eq:phi}.  We can similarly square the perturbation and average over orbital phase and population factors to get
\begin{eqnarray}
\sigma^2_\phi =   \Var(\phi) & \approx & \Var(\tau) + 2\Var(e_x) + 2\Var(e_y) \\
             & \approx & 6T\nIr + \left(8T+3T^3\right) \nIphi.
                         \label{eq:varphi}
\end{eqnarray}
The heliocentric latitude (or equivalently, polar angle) has a deviation
\begin{eqnarray}
  \Delta\theta & = & -\Delta L_x \frac{x}{r} - \Delta L_y \frac{y}{r} \\
  \Rightarrow \quad \sigma^2_\theta = \Var(\theta) & \approx & \langle\cos^2\tau\rangle \Var(L_x) + \langle\sin^2\tau\rangle \Var(L_y) \\
               & \approx & \frac{T}{2} \nIz.
                           \label{eq:vartheta}
\end{eqnarray}

>>>>

\begin{abstract}
  The explosion of high-precision astrometric data on main-belt
  asteroids (MBAs) enables new
  inferences on gravitational and non-gravitational forces present in
  this region.  We estimate the size of MBA motions
  caused by mutual gravitational encounters with other MBAs that are
  either omitted from ephemeris models or have uncertain mass
  estimates. In other words, what is the typical Brownian motion among
  MBAs that cannot be predicted from the ephemeris, and therefore
  serves as noise on inferences from the MBAs?  We estimate the RMS
  azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by
  numerical estimation of the distribution of impulse sizes and
  directions among known MBAs, combined with analytical propagation
  into future positional uncertainties. At current
  levels of asteroid-mass knowledge, $\sigma_\phi$ rises to
  $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
  $T^{3/2},$  large enough to degrade many inferences from
  Gaia and LSST MBA data. LSST data will, however, improve MBA
  mass knowledge enough to lower this Brownian uncertainty by
  $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are
  factors $\approx7$ and $\approx45,$ respectively,
  lower than the azimuthal noise, and grow as $T^{3/2}$
  and $T^{1/2}.$  For full exploitation of Gaia and
  LSST MBA data, ephemeris models should include the
  $\approx1000$~largest asteroids as active bodies with free masses,
  even if not all are well constrained.  This will correctly propagate
  the uncertainties from these 1000 sources' deflections into desired
  inferences. The RMS value of deflections from less-massive MBAs is
  then just $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

\keywords{Main belt asteroids (2036), Asteroid dynamics (2210), Astrometry (80)}

@@

We now have all the information we need to create the $\matA$ matrix.
Rotating to $\vecq=\{a, \tau, e_x, e_y, L_x, L_y\}$ and with $\vecI=\{I_r, I_\phi, I_z\}$ we have, to $O(e)$,
\begin{equation}
         \matA(e,T,t_i) = \left( \begin{array}{ccc}
 2e\sin\tau & 2 + 2e\cos\tau & 0 \\
 -2+e\left[1.5\cos\tau-3(T-t_i)\sin\tau\right] & e\sin\tau -3(1+e\cos\tau)(T-t_i) & 0\\
 \sin\tau+e\sin 2\tau & 2\cos\tau - 3e\sin^2\tau & 0\\
  -\cos\tau - 2e\sin^2\tau & 2\sin\tau-1.5e\sin 2\tau & 0 \\
0 & 0 & \sin\tau + 0.5 e \sin 2\tau \\
 0 & 0 &  -\cos\tau +e(3-\cos 2\tau)/2 \\
 \end{array}\right)
\end{equation}
We can now do the time integral and sum over impulse directions in \eqq{eq:Cqjk}, and average over $\tau$ to average over the orbital phase of the impulses. The diagonal elements of $\covm^q$ are
\begin{eqnarray}
  \Var(a) & = & 4T \nIphi \nonumber\\
  \Var(\tau) & = & 4T\nIr + 3T^3\nIphi \nonumber\\
  \Var(e_x) =  \Var(e_y) & = & \nIr/2+2\nIphi \nonumber\\
  \Var( L_x) =  \Var( L_y) & = & \nIz/2
\label{eq:Cq}
\end{eqnarray}
All of the off-diagonal terms (covariances) are either zero or $O\left(e\left\langle nI^2\right\rangle\right),$ and they result in observable consequences that are smaller by $\left\langle e^2 \right\rangle$ than the diagonal terms' contributions.

 \subsection{Astrometric and ranging shifts from elements shifts}
\label{sec:observe}
The radial coordinate of a Keplerian orbit is $r=a(1-e\cos E),$ and an expansion of the eccentric anomaly $E$ in powers of $e$ is
\begin{equation}
  \label{eq:Ee2}
  e \cos E  =  e\cos \ma - (e \sin \ma)^2 + O(e^3).
\end{equation}
From this we can derive the perturbations to $r$ as a function of our chosen basis for perturbations to the orbital elements.  Making use of Equations~(\ref{eq:ecosM}) and (\ref{eq:esinM}) we obtain
\begin{equation}
  \Delta r = (1-e\cos\tau) \Delta a + e\sin\tau \Delta\tau +  (-\cos\tau -2e\sin^2\tau) \Delta e_x + (-\sin\tau+e\sin 2\tau) \Delta e_y.
\end{equation}
If we square this expression, average over the orbital phase $\tau$ of the observation, and retain leading order in $e$, we get the variance of the radial coordinate from the unperturbed orbit.  All of the terms involving covariances between orbital elements either average to zero over an orbit, or are suppressed by $e^2$ relative to other terms.  We are left with
\begin{eqnarray}
  \sigma^2_r = \Var(r) & \approx & \Var(a) + \frac{1}{2}\Var(e_x) + \frac{1}{2}\Var(e_y) + \frac{\langle e^2\rangle}{2} \Var(\tau) \\
          & \approx & \frac{T}{2} \nIr + \left( 6T + \frac{3\langle e^2 \rangle}{2} T^3\right) \nIphi.
                      \label{eq:varr}
\end{eqnarray}
The second line substitutes the results from Equations~(\ref{eq:Cq}).  We retain the final term because, as $T>>1,$ this precession term that scales with $\langle e^2 \rangle T^3$ becomes comparable to or larger than the other terms that scale as $T$.  
The heliocentric azimuthal angle $\phi$ is given by \eqq{eq:phi}.  We can similarly square the perturbation and average over orbital phase and population factors to get
\begin{eqnarray}
\sigma^2_\phi =   \Var(\phi) & \approx & \Var(\tau) + 2\Var(e_x) + 2\Var(e_y) \\
             & \approx & 6T\nIr + \left(8T+3T^3\right) \nIphi.
                         \label{eq:varphi}
\end{eqnarray}
The heliocentric latitude (or equivalently, polar angle) has a deviation
\begin{eqnarray}
  \Delta\theta & = & -\Delta L_x \frac{x}{r} - \Delta L_y \frac{y}{r} \\
  \Rightarrow \quad \sigma^2_\theta = \Var(\theta) & \approx & \langle\cos^2\tau\rangle \Var(L_x) + \langle\sin^2\tau\rangle \Var(L_y) \\
               & \approx & \frac{T}{2} \nIz.
                           \label{eq:vartheta}
\end{eqnarray}
